---
title: "My markdown"
author: "Boaz"
date: "2024-11-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load packages
```{r}
library(tidyverse)
library(viridis)
library(plotly)
```

#Loading the data

"The dataset is loaded directly from my week 8 Github repository, and it contains information about the number of federated sports licenses held by men and women in different sports."

```{r}
adm_data <- read_csv("alcohol_related_admissions.csv")
```

Data inspection

```{r}
head(adm_data, 10)
glimpse(adm_data)
str(adm_data)


table(adm_data$year)
```
#Data preparation

"The data is prepared by calculating the percentage difference between the number of men and women holding federated licenses in each sport. The data is then reshaped into a long format for easier plotting."

Let's compute means of admissions in each health board

```{r}
mean_admissions <- adm_data %>% group_by(year, health_board) %>%
    summarise(mean_admission = mean(alcohol_admissions))

mean_admissions   
```

#Data visualisation

Let's plot a simple bargraph

```{r}
mean_admissions %>% filter(year == 2019) %>% 
    rename(alcohol_admissions = mean_admission) %>% 
    ggplot(aes(health_board, alcohol_admissions)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    ylab("Age-sex standardised rate per 100,000") +
    xlab(NULL) +
    ggtitle("Alcohol-related hospital admissions in 2019")
```
let's reorder scores
```{r}
mean_admissions %>% reorder(health_board, alcohol)

mean_admissions %>% filter(year == 2019) %>% 
    rename(alcohol_admissions = mean_admission) %>% 
    ggplot(aes(reorder(health_board, alcohol_admissions), alcohol_admissions)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    ylab("Age-sex standardised rate per 100,000") +
    xlab(NULL) +
    ggtitle("Alcohol-related hospital admissions in 2019")
```
Exercise 1: Could you make Grampian stand out by giving it a different colour?
```{r}
mean_admissions %>% filter(year == 2019) %>% 
    rename(alcohol_admissions = mean_admission) %>% 
    ggplot(aes(reorder(health_board, alcohol_admissions), alcohol_admissions, 
               fill = health_board == "Grampian")) +
    geom_bar(stat = "identity") +
    coord_flip() +
    ylab("Age-sex standardised rate per 100,000") +
    xlab(NULL) +
    ggtitle("Alcohol-related hospital admissions in 2019")
```

Let's make a lollipop graph instead
```{r}

```





