---
title: "week 6"
author: "Boaz"
date: "2024-11-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages
```{r load packages}
library(tidyverse)
library(dplyr)
```

Let's read in data
```{r}
prom <- read.csv("Hip Replacement CCG 1819.csv")
glimpse(prom)
```

Exercise 1:
Now we select variables of interest for this practical
```{r}
eq5d <- prom %>% select("Gender", "Pre.Op.Q.EQ5D.Index", "Post.Op.Q.EQ5D.Index")
head(eq5d, 10)

```

Data cleaning: 

rename colomns

```{r}
eq5d <- eq5d %>% rename(Pre_op_eq5d_index = "Pre.Op.Q.EQ5D.Index", 
                Post_op_eq5d_index = "Post.Op.Q.EQ5D.Index" )
```


The colomn Gender have some * which does not make sense, I prefer to remove those rows
```{r}
eq5d$Gender %>% unique()
eq5d$Gender %>% table()

eq5d_gender_nona <- eq5d %>% filter(Gender != "*")

head(eq5d_gender_nona, 20)
```

Let's recode this
```{r}
eq5d_gender_nona <- eq5d_gender_nona %>% 
    mutate(Gender = ifelse(Gender == 1, "Male", "Female"))


head(eq5d_gender_nona, 20)

```

tidy data
```{r}
eq5d_gender_nona_longer <- eq5d_gender_nona %>% pivot_longer(c(Pre_op_eq5d_index, Post_op_eq5d_index),
                                  names_to = "Time",
                                  values_to = "Q.eq5d")
eq5d_gender_nona_longer <- eq5d_gender_nona_longer %>% 
    mutate(Time = ifelse(Time == "Pre_op_eq5d_index", "Pre_op", "Post_op"))

eq5d_gender_nona_longer$Time <- factor(eq5d_gender_nona_longer$Time, levels = c("Pre_op", "Post_op"))

head(eq5d_gender_nona_longer, 10)
```


Plot 'EQ-5D Index' scores pre and post operation for each gender
```{r}
eq5d_gender_nona_longer %>% drop_na(Q.eq5d) %>% 
    ggplot(aes(Gender)) +
    geom_bar(aes(fill = Gender))

eq5d_gender_nona_longer %>% drop_na(Q.eq5d) %>% 
    ggplot(aes(Time, Q.eq5d)) +
    geom_col(aes(fill = Gender), width = 0.8) + 
    theme_minimal()
    
eq5d_gender_nona_longer %>% drop_na(Q.eq5d) %>% 
    ggplot(aes(Time, Q.eq5d)) +
    geom_boxplot(aes(fill = Gender), width = 0.8) + 
    theme_minimal()

eq5d_gender_nona_longer %>% drop_na(Q.eq5d) %>% 
    ggplot(aes(Time, Q.eq5d)) +
    geom_boxplot(width = 0.8) + 
    theme_minimal()
```
Assumption check 
Summary stats and histogram
```{r}
summary_stat <- eq5d_gender_nona_longer %>% drop_na(Q.eq5d) %>% 
    group_by(Time) %>% summarise(mean(Q.eq5d), 
                                 sd(Q.eq5d), 
                                 quantile(Q.eq5d, probs = c(0.25)),
                                 median(Q.eq5d),
                                 quantile(Q.eq5d, probs = c(0.75)),
                                 )
summary_stat

gather(summary_stat, key = "Time", value = "value", "mean(Q.eq5d)","sd(Q.eq5d)", 
                                 "quantile(Q.eq5d, probs = c(0.25))",
                                 "median(Q.eq5d)",
                                 "quantile(Q.eq5d, probs = c(0.75))")


eq5d_gender_nona$d <-  eq5d_gender_nona$Post_op_eq5d_index - eq5d_gender_nona$Pre_op_eq5d_index
head(eq5d_gender_nona)

# histogram
eq5d_gender_nona %>% drop_na(d) %>% 
    ggplot(aes(d)) +
    geom_histogram()
```

The appropriate statistic test is the Wilcoxon test

H0 : There is no change in quality of life after hip operation or There is no difference in distribution of differences between quality of life before and after hip operation.
```{r}

wilcox.test(eq5d_gender_nona$Pre_op_eq5d_index, eq5d_gender_nona$Post_op_eq5d_index, paired = TRUE, conf.int = 0.95)


```


