---
title: "Week 7"
author: "Boaz"
date: "2024-11-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load package
```{r}
library(tidyverse)
```

Let's read in data
```{r}
deprivation <- read_csv("alcohol-admissions_deprivation-data.csv")
head(deprivation, 30)
```

Inspect the data
```{r}
glimpse(deprivation)
table(deprivation$quintile)
```

Exercise 1: For the alcohol-related hospital admission data per intermediate data zone, plot the data distributions for each health board in 2019.

Let's remove financial year in the colomn "period"
```{r}
deprivation <- deprivation %>% mutate(period = gsub(" financial year", "", period))
deprivation
```



Let's get the year of interest (2019)
```{r}
table(deprivation$period)

deprivation %>% filter(period == "2019/20") %>% 
    ggplot(aes(indicator_measure)) +
    geom_histogram(bins = 15) +
    facet_wrap(~geography_code)

dep2019 <- deprivation[which(deprivation$period == "2019/20"), ] 

dep2019 %>% ggplot(aes(geography_code, indicator_measure)) +
    geom_boxplot(aes(fill = geography_code))
```
Exercise 2: For which health boards do we have data on alcohol-related hospital admissions for each deprivation quintile? (Hint: you can use the same geography dataset with intermediate zones that we used in the previous section to find the health board names.
```{r}
table(deprivation$geography_code)

deprivation$heathboard <- factor(deprivation$geography_code, levels = c("S08000020", "S08000029"), 
                                 labels = c("NHS Grampian", "NHS Fife"))
deprivation
```

Exercise 3: Create boxplots showing alcohol admissions for each deprivation quintile. Create a different plot for each health board in the data set. 
```{r}
deprivation %>% ggplot(aes(quintile, indicator_measure)) +
    geom_boxplot(aes(fill = quintile))

deprivation %>% ggplot(aes(heathboard, indicator_measure)) +
    geom_boxplot(aes(fill = heathboard))
```

